---
title: "Cousera Reproducible research week 2"
author: "Edwin Wanner"
date: "9 oktober 2018"
output: html_document
html_document:
keep_md: true
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(dev = c('png'), fig.path = 'figure/',echo = TRUE)
```

## Reproducible research - Project 1
This R Markdown document is the report for the Coursera course Reproducible research - week 2.
A simple set of data analysis tasks will be performed.

First, reading in the data:

```{r data}
steps <- read.csv("activity.csv")
```

Next, we will transform the date column to be of the type 'date'.

```{r transform}
steps$date <- as.Date(steps$date)
```

## Mean total number of steps per day
In the following tasks, we will ignore the NA.

Calculating the total number of steps taken per day:

```{r steps per day}
stepsPerDay <- aggregate(steps ~ date,data = steps, FUN = sum, na.action = na.omit )
stepsPerDay
```

The table shows that that there are missing dates. These are left out on purpose.

To gain more insight into the data, we will create a histogram displaying the total number of steps for each day.

```{r histogram steps}
hist(stepsPerDay$steps,
     main = "Number of steps per day",
     xlab = "Number of steps", 
     breaks = 10, 
     col = "grey",
     xlim = c(0,25000))

```

Now that we know the total number of steps per day, we are now interested in the mean and median number of steps per day.

```{r mean median steps}
stepsMean   <- aggregate(steps ~ date,data = steps, FUN = mean, na.action = na.omit )
stepsMedian <- aggregate(steps ~ date,data = steps, FUN = median, na.action = na.omit )
merged <- merge(stepsMean, stepsMedian, by = "date")
names(merged) <- c("date","mean","median")
merged

```

The table clearly shows that the mean is known everywhere, though the median is zero.

##Average daily activity pattern
We will dive deeper into the number of steps, by looking at the 5-minute intervals.

```{r 5-minute}
stepsPerInterval <- aggregate(steps~interval, data = steps, FUN = mean, na.action = na.omit)
with(stepsPerInterval, plot(interval, steps, type = "l", col = "red", xlab = "5-minute interval", main = "Average number of steps per 5-minute interval"))

maxInterval <- stepsPerInterval[which.max(stepsPerInterval$steps),1]
abline(v= maxInterval, lwd = 3, lty = 2)
```

The interval that has, on average, the highest number of steps is `r maxInterval`.

## Imputing missing values
We know that there are missing values in the data set.

```{r NA}
missing <- sum(is.na(steps$steps))
```
The total number of NA values in the data set is `r missing`.

In order to fill in the missing values, we will use the average 5-minute intervals to replace the missing values.

```{r NA replace}
idNA <- which(is.na(steps$steps))
stepsFilled <- steps
for( i in 1:length(idNA)) {
        stepsFilled[idNA[i],1] <- stepsPerInterval[which(stepsPerInterval$interval == steps[idNA[i],3]),2]        
}
```

Using the dataset without the missing values, the histogram of the total number of steps will look as follows:

```{r histogram missing}
stepsPerDayFilled <- aggregate(steps ~ date,data = stepsFilled, FUN = sum, na.action = na.omit )
hist(stepsPerDayFilled$steps,
     main = "Number of steps per day",
     xlab = "Number of steps", 
     breaks = 10, 
     col = "grey",
     xlim = c(0,25000))
```

The mean and median are then as follows:

```{r mean median missing}
stepsFilledMean   <- aggregate(steps ~ date,data = stepsFilled, FUN = mean, na.action = na.omit )
stepsFilledMedian <- aggregate(steps ~ date,data = stepsFilled, FUN = median, na.action = na.omit )
mergedFilled <- merge(stepsFilledMean, stepsFilledMedian, by = "date")
names(mergedFilled) <- c("date","mean","median")
mergedFilled
```

We can clearly see some differences in the median, which is not always zero. Furthermore, the dates that had a 'NA' value, are now shown, which were ignored in the previous table. The impact of of filling in the missing values is that all days are now shown and not only a selection. The pattern in the histogram is similar, but the frequency is higher.

## differences in activity pattern between weekdays and weekends
There might be a difference in the activity pattern between weekdays and the weekends. In order to investigate this, a new variable will be created that displays whether the given date is a weekday or a weekend. 

```{r weekend}
Sys.setlocale("LC_ALL","English")
stepsFilled$Weekday <- weekdays(stepsFilled$date)
stepsFilled$daytype <- as.factor(ifelse(stepsFilled$Weekday %in% c("Saturday","Sunday"), "Weekend","Weekday"))
```

The panel plot below shows a time series plot of the 5-minute interval average number of steps on weekdays and weekends.

```{r histogram weekend}
library(ggplot2)
par(mfrow = c(1,2))
stepsIntervalDaytype <- aggregate(steps~interval+daytype, data = stepsFilled, FUN = mean, na.action = na.omit)
ggplot(stepsIntervalDaytype, aes(interval, steps,daytype)) +
        geom_line() +
        facet_wrap(~daytype, ncol =1) +
        labs(y = "number of steps", x = "5-minute interval")
```

The plot for the weekday shows a high peek at the start of the day. In the weekends the average number of steps has more peeks.